---
title: "MatchItEXT: Supplementary to MatchIt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MatchItEXT: Supplementary to MatchIt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

MatchItEXT is a supplementary package to MatchIt. As the name implies, it can 
be regarded as an extension of MatchIt. It draws on the results from the latter 
to do further computation or plotting and to generate several outputs that are 
not available in MatchIt. A researcher might be interested in these outputs 
because they can help diagnose the matching result.

For now, functions in MatchItEXT (0.1.0) mainly include the following three 
aspects:

1. Computing Standardized Mean Difference (SMD).
1. Drawing two types of comparison plot before and after matching to show the 
matching result.
1. Computing two types of variance ratio as diagnostics of the matching result.

This tutorial will use the data set _lalonde_ in MatchIt to demonstrate how to 
apply MatchItEXT functions. The readers of this tutorial are assumed to have 
used MatchIt package before. Note the analyses below are only for the purpose of 
demonstration, and without a solid theoretical foundation. Therefore, the result
 of matching in this tutorial is not of substantial meaning. The interpretation 
of results is omitted.

###### Step 1: Loading the required packages and data set
First of all, if the required packages are not installed yet, we need install 
them first. If they have been installed, this step can be skipped.
```{r install, eval= FALSE}
pkg <- c("MatchIt", "MatchItEXT", "ggplot2")
install.packages(pkg)
```

Then, we can load required packages and example data set _lalonde_. If you are 
not familiar with this data set, please refer to the documentation of MatchIt 
package 
([click here](https://cran.r-project.org/web/packages/MatchIt/MatchIt.pdf)).
Note there are different versions of _lalonde_ data set online, the sample size 
and variable names might slightly differ from each other. 

```{r setup}
library(MatchIt)
library(MatchItEXT)
library(ggplot2)
data(lalonde)
```

###### Step 2: Checking data set lalonde
Since the data set has been loaded, we can have a glance at it.
```{r}
## Checking the size and structure of lalonde.
str(lalonde)

## Checking the first 5 rows of lalonde.
head(lalonde, 5)

## Checking whether there is any missing value in each column.
sapply(lalonde, function(x) sum(is.na(x)))

## No missing value was found.
```
###### Step 3: Generating a MatchIt result
To use the functions in MatchItEXT, most of the time we need results from 
MatchIt. So we would first generate an example from matchit() for later use. if 
you are new to MatchIt, please read this [official tutorial](https://imai.fas.harvard.edu/research/files/matchit.pdf).
```{r}
## Noticing that dichotomous variables that would be used  are of numeric type, 
## they should be converted to factor type beforehand.
lalonde$treat <- as.factor(lalonde$treat)
lalonde$black <- as.factor(lalonde$black)
lalonde$hispan <- as.factor(lalonde$hispan)
## Creating a propensity score estimation formula.
## In the formula, treat, black, and hispan are treated as binary variables.
formu <- as.formula(treat ~ re74 + re75 + age + educ + black + hispan)
## Applying Nearest neighbor matching on the selected covariates. The result 
## would be used as an input in the MatchItEXT functions. You can try other 
## matching methods such as "optimal", "full, "genetic", etc.
m_near <- MatchIt::matchit(formula = formu, data = lalonde, distance = "logit",
                           method = "nearest")
```
###### Step 4: Computing the Standardized Mean Difference (SMD)
SMD is defined as the difference between the focal and reference group divided 
by a standard deviation (Austin, 2011). If we compute the SMD before and after 
matching, we would know whether the SMD is broadened or narrowed after matching.
 Thus, it can be used as a diagnostic of matching result. An absolute value of 
SMD less than 0.1 is considered as an acceptable value that shows enough 
similarity between groups (Austin, 2011). MatchItEXT applies formulae provided 
by Austin (2011) to compute SMD. Note the formula for continuous variables 
differs from that for binary variables. For multiple categorical variables, they
 can be represented by a bunch of binary variables (Austin, 2011). 

As for the computation of standard deviation (SD), usually there are two options
. One is a simple pooled SD based on both treatment group and control group 
before matching. The other is just the SD of original treatment group. N.B. that
 MatchIt package uses the latter (Ho et al., 2011). Thus, there is a _sd_ 
argument in the function, if you want to be consistent with MatchIt, you can 
choose "treatment". The default is "pooled". Also note the SMD can be larger 
than one, which just means the difference is larger than one SD. 
```{r}
## Computing Standardized Mean Difference (SMD).
## If the matching result is satisfying, most SMD after matching would drop in 
## the range of -0.1 ~ 0.1, see the column "smd_after".
smd_near <- compute_smd(mi_obj = m_near, sd = "pooled")
smd_near

## Column names explanation:
## mean_tr_bf: mean of treatment group before matching
## mean_ctl_bf: mean of control group before matching
## mean_tr_af: mean of treatment group after matching
## mean_ctl_af: mean of control group after matching
## var_type: variable type
## var_tr_bf: variance of treatment group before matching
## var_ctl_bf: variance of control group before matching
## sd_bf_pooled: pooled standard deviation before matching
## sd_bf_tr: standard deviation of treatment group before matching
## smd_before: SMD before matching
## smd_after: SMD after matching

## Switching "pooled" SD to "treatment" SD.
smd_near <- compute_smd(mi_obj = m_near, sd = "treatment")
smd_near
```
From the above results, the first row "distance" denotes the overall distance 
measure (i.e., the propensity score in our case). It follows a set of covariates
 used in the formula. The last two columns in the data frame are SMD before and 
after matching.

MatchIt package supports various matching methods. As of version 3.0.2, it can 
apply six methods, including "exact", "full", "genetic", "nearest", "optimal", 
and "subclass". Among these methods, the result of "full", "genetic", "nearest",
 and "optimal" are applicable to compute_smd(), while the result of "exact" is 
not applicable. For the result of "subclass", we can apply compute_sub_smd(), as
 shown below.
```{r}
## Generating a subclassification result from matchit() with the same formula.
m_sub <- MatchIt::matchit(formula = formu, data = lalonde, distance = "logit",
                           method = "subclass", subclass = 5)
smd_sub <- compute_sub_smd(mi_obj = m_sub, sd = "pooled")
## The threshold of good matching for SMD is ± 0.1. The closer to 0, the better.
smd_sub
```
Different from compute_smd(), the result of compute_sub_smd() only provides a 
single SMD, indicating the overall difference of distance measure after 
subclassification. There is no separate SMD for each covariate when using 
subclassification method.

###### Step 5: Drawing SMD comparison dot-and-line plot
Although MatchIt provides a function to draw similar dot-and-line comparison 
plot (e.g., plot(summary(m_near, standardize = TRUE))), the following plot 
function differs from its counterpart in that: 

1. coloring different types of variable: SMD of overall distance is colored in 
blue, increased SMD after matching is colored in red, unchanged SMD is colored 
in green, and decreased SMD is colored in gray.
1. using original SMD rather than the absolute value of SMD. Hence, there would 
be negative SMD.
1. drawing horizontal lines of -0.2, -0.1, 0.1, and 0.2 to help distinguish 
matching result areas.
1. providing R code for further revision if the user is not satisfied with the 
plot.

```{r, fig.width = 7, fig.height = 5}
## If the matching result is satisfying, most SMD after matching would rest in 
## the range of -0.1 ~ 0.1.
plot_smd_near <- plot_smd(smd_near)
plot_smd_near$plot
## Increased SMD are stored in the returning result, R code for plotting and 
## other relevant data are available as well, use '$' to obtain them.
plot_smd_near$smd_increase
plot_smd_near$plot_code
```
###### Step 6: Drawing QQ plot for group comparison on the distance measure
Although MatchIt provides a function to draw QQ plots (e.g., plot(m_near), the 
default plot type for MatchIt results), it only generates QQ plots for 
covariates, not for the overall distance measure. The following function 
plot_ps_qq() can draw it for you.
```{r, fig.width = 7, fig.height = 5}
## QQ plot for overall distance measure (propensity score in our case).
## If the matching result is good enough, most points in the QQ plot after 
## matching would rest on the 45-degree line.
m_near_qq <- plot_ps_qq(m_near)
m_near_qq$plot
```

###### Step 7: Computing variance ratio and residual variance ratio
Besides SMD, Rubin (2001) proposed two other diagnostics for propensity score 
matching, the one is the ratio of the variances of the propensity scores 
between the two groups (treatment group to control group), the other is the 
ratio of the variance of the residuals orthogonal to the propensity scores 
between the two groups, for each of the covariates. As for the criterion, the 
ideal ratio should be close to one, which means variances from two groups are 
similar. A ratio of less than 1/2 or larger than 2 is far too extreme (Rubin, 
2001). These two diagnostics are less popular in practice. We can compute both 
with the following functions.
```{r}
## Computing variance ratio of propensity score
m_near_var_ratio <- compute_var_ratio(m_near)
m_near_var_ratio

## Column names explanation:
## var_tr_bf: variance of treatment group before matching
## var_ctl_bf: variance of control group before matching
## var_tr_af: variance of treatment group after matching
## var_ctl_af: variance of control group after matching
## ratio_bf: variance ratio before matching
## ratio_af: variance ratio after matching

## Computing residual variance ratio for each covariate
## Checking the formula for propensity score estimation
parse_formula(m_near)
## This auxiliary function shows covariates in the formula to help decide 
## variable types.

## The function requires the original data set, the MatchIt object (except for 
## results from "exact" or "subclass" method), and a vector specifying 
## covariate types.
## Valid values for covariate types:
## do not include this covariate: 'excluded', '0', 0, NA;
## continuous variable: 'continuous', '1', 1;
## dichotomous variable: 'binary', '2', 2;
## ordinal variable: 'ordinal', '3', 3.
## Note a vector in R does not support different types of values, thus these 
## valid values cannot be mixed.

## Note multinomial variable is not applicable to this function.

## The 'discard' argument is logical to judge if some cases were discarded 
## before matching. The default is FALSE, no discard before matching. 
compute_res_var_ratio(original_data = lalonde, mi_obj = m_near, type_vec = 
                        c(0, 1, 1, 1, 2, 2), discard = FALSE)
```


#### Summary
As shown above, MatchItEXT package helps diagnose the result of matching by 
calculating SMD, variance ratio and residual variance ratio. Besides that, it 
can draw SMD dot-and-line plot and QQ plot to compare the matching result before
 and after matching. These are unavailable in MatchIt package. Therefore, it is 
a supplementary package to MatchIt.


#### References
Austin, P. C. (2011). An Introduction to Propensity Score Methods for Reducing 
the Effects of Confounding in Observational Studies. 
_Multivariate Behavioral Research, 46_(3), 399–424. https://doi.org/10.1080/00273171.2011.568786

Ho, D. E., Imai, K., King, G., & Stuart, E. A. (2011). MatchIt: Nonparametric 
Preprocessing for Parametric Causal Inference. 
_Journal of Statistical Software, 42_(8). 
https://doi.org/10.18637/jss.v042.i08

Rubin, D. B. (2001). Using propensity scores to help design observational 
studies: Application to the tobacco litigation. 
_Health Services and Outcomes Research Methodology, 2_(3/4), 169–188. https://doi.org/10.1023/A:1020363010465
